---
title: "Cochineal 16S - Preliminary Results"
author:
  - name: João Vitor Cavalcante
    orcid: 0000-0001-7513-7376
  - name: Bianca Santiago
    orcid: 0000-0002-8037-6264
  - name: Rodrigo Dalmolin
    orcid: 0000-0002-1688-6155
format: html
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = TRUE
)
```

```{r}
library(vroom)
library(dplyr)
library(here)
library(DT)
library(stringr)
library(reactable)
library(ggplot2)
library(pheatmap)
library(tidyr)

make_datatable <- function(data, col_names) {
  data %>%
    datatable(
      .,
      extensions = "Buttons",
      rownames = FALSE,
      escape = FALSE,
      colnames = col_names,
      filter = 'top',
      options = list(
        lengthMenu = c(5, 10, 25, 50, 100),
        pageLength = 5,
        scrollY = 200,
        scroller = TRUE,
        dom = '<"d-flex justify-content-between"lBf>rtip',
        buttons = list(
          'copy',
          list(
            extend = 'collection',
            buttons = c('csv', 'excel'),
            text = 'Download'
          )
        )
      )
    )
}

samples <- c(
  "GRUPO_1_1",
  "GRUPO_1_2",
  "GRUPO_1_3",
  "GRUPO_2_1",
  "GRUPO_2_2",
  "GRUPO_2_3",
  "GRUPO_3_1",
  "GRUPO_3_2",
  "GRUPO_3_3"
)
```

# Methodology

Data was processed with version 2.10.0 of the nf-core/ampliseq pipeline [@straub2020]. The pipeline uses DADA2 [@callahan2016] for Amplicon-sequence variant (ASV) inference, QIIME2 [@bolyen2019] for calculating diversity indices and PICRUSt2 [@douglas2020] for predicting the functional potential of bacteria.

# Results

## Relative Taxonomic Abundance

The full data for the calculated taxonomic abundance can be interacted with in the following page: <https://dalmolingroup.github.io/cochineal_16s/qiime2_relative_abundance/>. It enables analysis at every taxonomic level, showing the relative frequency of each taxa at that level.

By far the most abundant taxa is *Candidatus Uzinura*, which has been described previously as an endosymbiont in armored scaled insects [@sabree2013].

## Alpha Diversity

The Shannon Diversity Index was calculated for each sample, here presented by the group they were assigned to in the sample name.

```{r}
alpha_div <-
  vroom(
    here(
      "data/ampliseq_results/qiime2/diversity/alpha_diversity/shannon_vector/metadata.tsv"
    ),
    comment = "#"
  )

ggplot(alpha_div, aes(y = group, x = shannon_entropy)) +
  geom_boxplot() +
  labs(
    y = element_blank(),
    x = "Shannon Index"
  ) +
  theme_classic()
```

## Functional Prediction

Below are the 30 most common pathways in all samples, as predicted by PICRUSt2. Pathway data is from the MetaCyc database [@caspi2018]. Values are the log10 transformed predicted abundances from PICRUSt2.

```{r}
abundance <- vroom::vroom(here("data/ampliseq_results/picrust/METACYC_path_abun_unstrat_descrip.tsv"))

common_paths <- abundance |> 
  rowwise() %>%
  mutate(
    GRUPO_1 = sum(GRUPO_11, GRUPO_12, GRUPO_13),
    GRUPO_2 = sum(GRUPO_21, GRUPO_22, GRUPO_23),
    GRUPO_3 = sum(GRUPO_31, GRUPO_32, GRUPO_33)
  ) |>
  select(-matches(".*\\d{2}$")) |>
  mutate(total_sum = sum(c_across(starts_with("GRUPO_")), na.rm = TRUE)) %>%
  ungroup() |> 
  slice_max(total_sum, n = 30) |> 
  select(-c(pathway, total_sum)) |> 
  tibble::column_to_rownames("description") |>  
  as.matrix()

pheat <- pheatmap(
  log10(common_paths),
  border_color = "white",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = RColorBrewer::brewer.pal(7, "Reds"),
  na_col = "#fafafa",
  fontsize = 10
  # ,main = title
)
pheat
```

# Acknowledgements

We would like to thank NPAD/UFRN for computational resources expended.
